Analysis of Programming Students' Prompts to Identify Interactions Patterns with Generative AI

### 2024-09-20 CSEDU2025/Paper.tex
update

% \cite{wu24} found that ChatGPT-supported learning not only promoted students'
% self-regulation but also increased their self-efficacy.

\begin{table*}[htbp]
  \caption{Categories of Analysis}
      \begin{center}
      \begin{tabular}{|p{4cm}|p{9cm}|}
      \hline
      \textbf{Category} & \textbf{Description} \\
      \hline
      Debugging Help & * Prompts that seek help to identify, fix errors, or understand the provided code snippet. \\
      Conceptual Question & * Prompts that are more about understanding concepts or algorithms rather than specific code. \\
      Student Correction & * Prompts where the student corrects the bot. \\
      \hline
      Code Snippet & * Prompts that ask for a specific part of the code, like a function or a segment. \\
      Complete Solution & * Prompts that request an entire solution or a complete code snippet. \\
      Multiple Question & * Prompts where the user wants to solve a multiple-question exercise. \\
      \hline
      Language Change & * Prompts that request a change of idiom to convey a message more effectively. \\
      Uncategorized & * Prompts that do not fit into any of the above categories. \\
      \hline
      \end{tabular}
      \label{tab:categories}
      \end{center}
  \end{table*}

@@

The study was conducted in two phases: categorization and analysis. The
categorization phase involved classifying student messages into eight
categories. The table \ref{tab:categories} shows the categories of analysis and
their description. The initial categories were based on the work of
\cite{Ghimire24}, however, four new categories were added due to the need to
give more detail to the classification. Beside of that, the categories were
grouped to understand the impact on critical thinking.

The analysis phase involved to understand the way that the students interact
with CharlieBot. The students messages were collected from the logs of the bot
and categorized manually by the authors. A total of YYY messages, of the first
semester of 2024, were analyzed. The participants were students of the YYY
semester of the Computer Science at UC3M where Java programming is taught.

\section{Results and Discussion}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[scale=0.6]{img/figure1.png}
    \caption{Classification of messages.}
    \label{fig:graph1}
\end{figure}

@@

@@

Messages classified as conceptual questions represent around 32\% of all
messages. Regarding Metacognitive Knowledge, about these sequences of prompts
related with conceptual questions, one possible interpretation may be that the
students are aware they have not fully mastered a concept and are trying to
understand it, demonstrating a good level of self-awareness. In this sense, the
student realizes that the problem requires an understanding of concepts in order
to advance in their studies. In this way, the student applies an active learning
strategy. About Metacognitive Experience, several types of feelings may be
associated at this moment, such as actively reflecting on the answers,
uncertainty and/or confusion when trying to understand the concept, frustration
for not progressing, and satisfaction as their questions are answered in a
way that facilitates understanding. Regarding the type of metacognition related
to the Task, the student recognized the difficult and may be trying to achieve a
deeper understanding of the evolved concepts to later apply in practice. With
respect to Metacognitive Strategies, the student demonstrates a metacognitive
planning process when identifying their knowledge gaps. They may start with more
general questions to obtain an overview of the concept and then ask more
specific questions as they gain more clarity. Because the questions are in
sequence, the student probably can adequately monitor their progress.

% Spacing
Spacing, a well-established learning strategy, was examined by analyzing
the time intervals between users’ interactions with CharlieBot. Although
only 12\% of the conversations exhibited message delays exceeding 24 hours,
40\% involved response times longer than 60 minutes. These findings suggest
that students typically do not extend conversations across multiple days.
However, the data also reveal that a substantial number of conversations
feature one or more pauses exceeding an hour, a practice known to enhance
knowledge retention.

% Categories Time

>>>>

% \cite{wu24} found that ChatGPT-supported learning not only promoted students'
\begin{table*}[htbp]
  \caption{Categories - adapted from \cite{Ghimire24}}
  \begin{center}
    \renewcommand{\arraystretch}{1.5} % Increase the spacing between rows
    \begin{tabular}{|p{3,5cm}|p{11cm}|}
      \hline
      \textbf{Category} & \textbf{Description} \\
      \hline
      Debugging Help & Prompts that seek help to identify, fix errors, or understand the provided code snippet. \\
      Conceptual Question & Prompts that are more about understanding concepts than specific code. \\
      Student Correction & Prompts where the student corrects the bot. \\
      \hline
      Code Snippet & Prompts that ask for a specific part of the code, like a function or a segment. \\
      Complete Solution & Prompts that request an entire solution or a complete code snippet. \\
      Multiple Question & Prompts where the user wants to solve a multiple-question exercise (Quiz). \\
      \hline
      Language Change & Prompts that request a change of idiom. \\
      Uncategorized & Prompts that do not fit into any of the above categories. \\
      \hline
    \end{tabular}
    \label{tab:categories}
  \end{center}
\end{table*}

@@

The study was conducted in three phases: data collection, categorization, and
analysis. During the first semester of 2024, student interactions with
CharlieBot were collected as part of the Java programming course at UC3M.
The categorization phase involved classifying student messages into eight
categories. The table \ref{tab:categories} shows the categories of analysis and
their description. The initial categories were based on the work of
\cite{Ghimire24}, however, four new categories were added due to the need to
give more detail to the classification. The analysis phase involved to
understand the way that the students interact with CharlieBot.
\section{\uppercase{Results and Discussion}}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[scale=0.62]{img/figure1.png}
    \caption{Classification of messages.}
    \label{fig:graph1}
\end{figure}

\begin{figure*}[htbp]
  \centering
  \includegraphics[scale=0.52]{img/figure2.png}
  \caption{Examples of conversations.}
  \label{fig:graph2}
\end{figure*}

@@

% Debugging help
\subsubsection*{Sequence of Debugging Help}

Around 30.9\% of the messages were classified as Debugging Help. Besides of
that, we identified a pattern of a sequences of Debugging Help in the
conversations. These prompts typically start with a message requesting a
Complete Solution or a Code Snippet. After that, the students normally chance
from passive to more active learning strategy.

In terms of Metacognitive Knowledge, the student probably recognizes that they
do not have the necessary tools or knowledge to debug the code on their own.
Concerning the understanding of the task, they may know that debugging is not
just a matter of superficial corrections but requires understanding the
underlying logic. Thus, the student apparently believes that the sequence of
questions related to debugging will help them understand the codes.

In terms of Metacognitive Experiences, some feelings may arise at this moment,
such as frustration or uncertainty in the face of errors, confusion during the
process, and also growing confidence or satisfaction as they better understand
the code and possible corrections.

With respect to Metacognitive related with the objective of the Task, the
student is probably focused on understanding the code in practical terms,
identifying errors, and learning how to better debug in the future.

As for Metacognitive associated with the Strategy, the student has requested
external help and realizes that dividing the debugging task into several
questions can result in a better understanding of the code. They can monitor
their progress throughout the process, adjusting their questions as necessary,
evaluate the effectiveness of the responses received, and reflect on what was
learned during the process.

\begin{figure*}[htbp]
  \centering
  \includegraphics[scale=0.55]{img/figure3.png}
  \caption{Interaction times per categories.}
  \label{fig:graph3}
\end{figure*}

@@

Messages classified as conceptual questions represent around 32\% of all
messages. We have found a pattern of sequences of Conceptual Questions in some
conversations. Some conversations are entirely composed of messages classified
as Conceptual Questions. In other cases, the sequence of Conceptual questions
starts with a request for a Complete Solution or a Code Snippet, followed by
a progression of Conceptual Questions. In both cases, the student is actively
engaging with the bot to understand the concepts and compromised with more
deep learning.

Regarding Metacognitive Knowledge, these sequences of prompts classified as
Conceptual Question, can be interpreted that the students are aware they have
not fully mastered a concept and are trying to understand it, demonstrating a
good level of self-awareness. In this sense, the student realizes that the
problem requires an understanding of concepts in order to advance in their
studies. In this way, the student applies an active learning strategy.

About Metacognitive Experience, several types of feelings may be associated at
this moment, such as actively reflecting on the answers, uncertainty and/or
confusion when trying to understand the concept, frustration for not
progressing, and satisfaction as their questions are answered in a way that
facilitates understanding.

Regarding the type of Metacognition related to objective of the Task, the
students may recognize the difficult and trying to achieve a deeper
understanding of the evolved concepts to later apply in practice.

With respect to Metacognitive Strategies, the student demonstrates a sence of
planning to identify their knowledge gaps. They may start with more general
questions to obtain an overview of the concept and then ask more specific
questions as they gain more clarity. Because the questions are in sequence, the
student probably can adequately monitor their progress.

\subsection*{Lack of Interleaving}

% Interleaving
Interleaving, a technique that involves alternating the conversation among
different types of categories.

% Eglington, L. G., & Kang, S. H. (2017). Interleaved presentation benefits
% science category learning. Journal of Applied Research in Memory and Cognition,
% 6, 475-485

\subsection*{Spacing}

% Spacing
Spacing, a well-established learning strategy \cite{Carvalho20}, was examined
by analyzing the time intervals between users’ interactions with CharlieBot.
Although only 7.4\% of the conversations exhibited message delays exceeding 24
hours, 26.19\% involved response times longer than 60 minutes. These findings
suggest that students typically do not extend conversations across multiple days.
However, the data also reveal that some conversations had one or more pauses
exceeding an hour, a practice known to enhance knowledge retention.
